Staged Compilation and Generativity \thanks{The author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}

### 2021-09-22 efop_abstract/ext_abstract.tex
typo
typo again

We give some examples for interpreting constructions in the model. We present the
results up to isomorphism, with some simplifications. We assume now that
object-level morphisms are substitutions. We have $\Bool_0 : \U_0$ and $\Bool_1
: \U_1$.
\begin{itemize}
  \item A closed function $t : \Bool_1 \to \Bool_1$ becomes a metatheoretical
    function in $\mathbb{B} \to \mathbb{B}$.
  \item A closed function $t : \Bool_0 \to \Bool_0$ becomes a closed object-theoretic
    function in $\Bool \to \Bool$.
  \item A closed function $t : \Code\,\Bool_0 \to \Bool_1$ becomes a function
    which maps a $\Bool$ term in any context to $\mathbb{B}$, such that the
    function commutes with object-theoretic substitution. For example if we have
    a variable $x$, we can substitute it with any term before feeding it to the
    semantic $t$, and the result is the same. In fact, this means that $t$
    cannot depend on its term argument, hence $t$ is specified simply by a
    $\mathbb{B}$.
  \item
    A closed function $t : \Bool_1 \to \Code\,\Bool_0$ becomes simply a pair of
    closed $\Bool$ terms.
\end{itemize}
\noindent\textbf{Yoneda lemma.} The Yoneda lemma is a general statement which
restricts the way meta-level values can depend on object-level ones. First, note that any
object-level typing context $\Gamma$ can be mapped to a presheaf, by taking the
sets of parallel substitutions into $\Gamma$. This is the \emph{Yoneda
embedding} of $\Gamma$, denoted by $\msf{y}\Gamma$. The Yoneda lemma says
that we have the following isomorphism of sets:
\[
   (\msf{y}\,\Gamma \Rightarrow \Delta) \simeq |\Delta|\,\Gamma
\]
where $\Rightarrow$ means a natural transformation, and $|\Delta|\,\Gamma$
denotes the set that we get by evaluating the $\Delta$ presheaf at the
object-level $\Gamma$ context. From this, what we essentially get is that any
2LTT term $\Gamma \vdash t : A$, such that $\Gamma$ is essentially interpreted
as $\msf{y}\,\Gamma'$ for some $\Gamma'$, is interpreted as an element of
$|A|\Gamma'$. We call $\Gamma$ \emph{representable} is there is such $\Gamma'$.

@@

Simple type theories still work, if we only have weakening in their equational
theory. For the perspective of staging, this is fine, because in \emph{code
generation} we care about the intensional definition of programs, and we do not
want to equate $\beta$-reducts, since a primary use-case of staging is to
improve runtime performance, hence distinguish between possibly
$\beta$-convertible programs.

>>>>

We give some examples for interpreting constructions in the model. We present the
results up to isomorphism, with some simplifications. We assume now that
object-level morphisms are substitutions. We have $\Bool_0 : \U_0$ and $\Bool_1
: \U_1$.
\begin{itemize}
  \item A closed function $t : \Bool_1 \to \Bool_1$ becomes a metatheoretical
    function in $\mathbb{B} \to \mathbb{B}$.
  \item A closed function $t : \Bool_0 \to \Bool_0$ becomes a closed object-theoretic
    function in $\Bool \to \Bool$.
  \item A closed function $t : \Code\,\Bool_0 \to \Bool_1$ becomes a function
    which maps a $\Bool$ term in any context to $\mathbb{B}$, such that the
    function commutes with object-theoretic substitution. For example if we have
    a variable $x$, we can substitute it with any term before feeding it to the
    semantic $t$, and the result is the same. In fact, this means that $t$
    cannot depend on its term argument, hence $t$ is specified simply by a
    $\mathbb{B}$.
  \item
    A closed function $t : \Bool_1 \to \Code\,\Bool_0$ becomes simply a pair of
    closed $\Bool$ terms.
\end{itemize}
\noindent\textbf{Yoneda lemma.} The Yoneda lemma is a general statement which
restricts the way meta-level values can depend on object-level ones. First, note that any
object-level typing context $\Gamma$ can be mapped to a presheaf, by taking the
sets of parallel substitutions into $\Gamma$. This is the \emph{Yoneda
embedding} of $\Gamma$, denoted by $\msf{y}\Gamma$. The Yoneda lemma says
that we have the following isomorphism of sets:
\[
   (\msf{y}\,\Gamma \Rightarrow \Delta) \simeq |\Delta|\,\Gamma
\]
where $\Rightarrow$ means a natural transformation, and $|\Delta|\,\Gamma$
denotes the set that we get by evaluating the $\Delta$ presheaf at the
object-level $\Gamma$ context. From this, what we essentially get is that any
2LTT term $\Gamma \vdash t : A$, such that $\Gamma$ is essentially interpreted
as $\msf{y}\,\Gamma'$ for some $\Gamma'$, is interpreted as an element of
$|A|\Gamma'$. We call $\Gamma$ \emph{representable} if there is such $\Gamma'$.

@@

Simple type theories still work, if we only have weakening in their equational
theory. From the perspective of staging, this is fine, because in \emph{code
generation} we care about the intensional definition of programs, and we do not
want to equate $\beta$-reducts, since a primary use-case of staging is to
improve runtime performance, hence distinguish between possibly
$\beta$-convertible programs.
